### 2011-11-07 inst/doc/article.Rnw
Article and internals

\SweaveOpts{prefix.string=Figs/fig}
\setkeys{Gin}{width=\textwidth} %% <- change width of figures
% Try to get the R code from running into the margin
<<echo=FALSE>>=
options(width=87, digits=3, scipen=4)
@

@@

\begin{center}
\vspace*{1in}
\textbf{\Large QCL: the added value of correlation }

@@

\textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
(QCL) mapping a novel approach to detect genetic regulation of phenotypes in 
recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
allows association of phenotype correlation differences to genetic markers, in almost 
the same way as QTL mapping associates genetic markers with differences in expression.
We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
into regulation of (gene) expression. QCL information in combination with classical QTL 
information leads to a more detailed insight into the genetic architecture underlying 
complex traits. 
In two example we reconstruct part of the genetic architecture in S. Cerevisiae grown 
under ethanol conditions. The other example shows the reconstruction of the chain 
elongation pathway in A. Thaliana. In both examples we map the observe differences 
in correlation to genetic markers, and show the biological relevance of QCL peeks and 
hotspots in regulatory network reconstruction.

@@

Genetical genomics experiments have shown that gene expression variation can be mapped to 
genotypic variation (Jansen and Nap, 2001). Quantitative Trait Locus (QTL) mapping of a 
gene identifies regions in the genome for which different genotypes lead to differential 
expression level of the gene. This QTL mapping can be done on all molecular levels including, 
gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).
Recently different methods have been proposed that allow studying, in addition to the genes 
differentially expressed between conditions, the genes that are differentially coexpressed 
between conditions (Tesson et al. 2010, Kompass et al. 2011). Such differential co-expression 
is taken as another level of information from genomic data. This level of information can for 
example be used to expose transcriptional network rewiring (Langfelder et al. 2008). Differential 
co-expression has been observed in studies of different diseases, including between healthy and 
cancer tissue studies. In parallel, correlations between phenotypes are studied for various 
reasons e.g. discovering and detecting batch effects, marker assisted breeding and causal 
inference. The importance of identifying differences in correlation has been increasingly 
recognized and this led to the development of this algorithm to perform QCL analysis in 
experimental crosses.

@@

\textbf{\sffamily Algorithm: eQCL mapping}

@@

\emph{\sffamily Step 1 Choosing a start/seed probe}

@@

\emph{\sffamily Step 2 Single phenotype QCL}

@@

\emph{\sffamily Step 2a Significance testing}

The threshold for a significant difference in QCL is assessed by using permutations, in 
each round the link between genotype and phenotype is broken, by redistributing at random 
genotypes amongst individuals not allowing for duplicates. The QCL on these new 
?ndividuals?allows for estimation of the underlying null-distribution of QCL scores.

\emph{\sffamily Step 2b Create single trait QCL profiles}

Single trait QCL profiles can be obtained by setting a significance threshold.

@@

Using the user defined seed gene we can reconstruct the interaction network, we 
first do a single phenotype QCL scan (step 2), then using a threshold we select 
phenotypes which show a significant change in QCL.

@@

\emph{\sffamily Step 3a Visualizing the Phenotypes x Marker relations}

When plotting multiple QCL profiles as a heatmap, multiple QCL are seen mapping to the 
same genetic location. At these locations, we expect to find one (or more) of the genes 
to show a cis-QTL, this phenotype (or phenotypes) introduced genetic variation, which is 
seen also (and more pronounced) in a QTL scan.

\emph{\sffamily Step 3b Visualizing the Phenotypes x Phenotypes relations}

\emph{\sffamily Step 4 Network reconstruction}

>>>>

\SweaveOpts{prefix.string=fig}
\setkeys{Gin}{width=\textwidth} %% <- change width of figures
% Try to get the R code from running into the margin
<<echo=FALSE>>=
options(width=87, digits=3, scipen=4)
@

@@

\begin{center}
\vspace*{1in}
\textbf{\Large QCL: Added value of differential correlation }

@@

\textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
(QCL) mapping, a novel approach to detect genetic regulation of phenotypes in 
recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
allows association of phenotype correlation differences to genetic markers, in almost 
the same way as QTL mapping associates genetic markers with differences in expression.
We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
into regulation of (gene) expression. QCL information in combination with classical QTL 
information leads to a more detailed insight into the genetic architecture underlying 
complex traits. 
In two example we reconstruct part of the genetic architecture in S. Cerevisiae grown 
under ethanol conditions. The other example shows the reconstruction of the chain 
elongation pathway in A. Thaliana. In both examples we map the observe differences 
in correlation to genetic markers, and show the biological relevance of QCL peeks and 
hotspots in regulatory network reconstruction.

@@

Genetical genomics experiments have shown that gene expression variation can be mapped to 
genotypic variation (Jansen and Nap, 2001). Quantitative Trait Locus (QTL) mapping of a 
gene identifies regions in the genome for which different genotypes lead to differential 
expression level of the gene. This QTL mapping can be done on all molecular levels including, 
gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

Recently different methods have been proposed that allow studying, in addition to the genes 
differentially expressed between conditions, the genes that are differentially coexpressed 
between conditions (Tesson et al. 2010, Kompass et al. 2011). Such differential co-expression 
is taken as another level of information from genomic data. This level of information can for 
example be used to expose transcriptional network rewiring (Langfelder et al. 2008). Differential 
co-expression has been observed in studies of different diseases, including between healthy and 
cancer tissue studies. In parallel, correlations between phenotypes are studied for various 
reasons e.g. discovering and detecting batch effects, marker assisted breeding and causal 
inference. The importance of identifying differences in correlation has been increasingly 
recognized and this led to the development of this algorithm to perform QCL analysis in 
experimental crosses.

@@

\textbf{\sffamily Algorithm: QCL mapping}

@@

\emph{\sffamily Choosing a start/seed probe (Optional)}

@@

\emph{\sffamily Single phenotype QCL}

@@

\emph{\sffamily Significance testing}

The threshold for a significant difference in QCL is assessed by using permutations, in 
each round the link between genotype and phenotype is broken, by redistributing at random 
genotypes amongst individuals not allowing for duplicates. The QCL on these new 
individuals allows for estimation of the underlying null-distribution of QCL scores. 
See figures XX, and XX showing the score distribution in permuted data.

\emph{\sffamily Creating single trait QCL profiles}

After permutation single trait QCL profiles can be obtained by setting a qcl significance 
threshold. This threshold defines when a change in correlation is significant, and at 
each marker we can plot the count of number the other phenotypes that show a significant 
difference in correlation. Overplotting chromosome, QTL information, and phenotype location 
(if applicable) produces the single QCL plots seen in figures XX, and XX.

@@

If we find any significant QCL information we can try to reconstruct the interaction 
network of the phenotype. We first do a single phenotype QCL scan (step 2), then 
using a permutation determin a qcl threshold, and select phenotypes which show a 
significant change in QCL.

@@

\emph{\sffamily Visualizing the Phenotypes x Marker relations}

When plotting multiple QCL profiles as a heatmap, multiple QCL are seen mapping to the 
same genetic location. At these locations, we expect to find one (or more) of the genes 
to show a cis-QTL, this is because a QTL can induce a QCL, because variation of expression 
in the phenotype is causing other phenotypes to become uncorrelated, when assigned to 
groups determined by the QTL marker. This phenotype (or phenotypes) introduced genetic 
variation, which is seen also (and more pronounced) in a QTL scan.

\emph{\sffamily Visualizing the Phenotypes x Phenotypes relations}

The additional information obtained by QCL mapping is summarized in the 
Phenotypes x Phenotypes matrix. This shows the summed evidence of interaction 
between the phenotypes showing QCL transmitted through all the markers and the phenotype scanned.

\emph{\sffamily Network reconstruction}
